### 2018-04-26 contents/ap_paper.tex
Updates from ShareLaTeX

An alternative to FA is to use the generalized fractional anisotropy (GFA) image. Since FA shows low intensity in cross-fiber regions, it may adversely affect registration in those areas. GFA can properly characterize crossing-fiber intensities, and therefore may be a better candidate than FA. The GFA is computed from the SH coefficients of spherical harmonics based methods such as Q-ball, where the diffusion orientation distribution function (ODF) is denoted as  $\Psi(u)$ \cite{Tuch2004}, such that:
\begin{equation}
GFA = \sqrt{\frac{n \sum_{i=1}^{n}( \Psi(u_i) - \langle\Psi\rangle)^2} {(n-1) \sum_{i=1}^{n}\Psi(u_i)^2}}
\end{equation}

@@

Another promising candidate is the anisotropic power (AP) image \cite{DellAcqua2014}. AP have similar visual intensity to T$_1$, and can be computed from HARDI acquisitions without any acquisition modifications. Since AP are derived from the SH model, it is also capable of proper cross-fiber region characterization. AP images are derived from the even harmonic orders ($l$) of SH coefficients \cite{Descoteaux2006,Frank2002} and are defined as
\begin{equation}
AP = \sum_{l=2,4,...} \frac{1}{2l+1}\sum_{m=-l}^{l} \lvert a_{l,m}\lvert^2
\end{equation}
where l is the harmonic order, m is the l-th order coefficient index, a is the corresponding SH coefficient. The AP values are then normalized to non-negative scale by the function:
\begin{equation}
\ln( \frac{AP}{AP_{ref}})
\end{equation}
where $\text{AP}_{\text{ref}}$ is the normalization constant. The resulting image will closely match T$_1$ intensities, and can better capture contrasts in the gray matter structures. AP values stabilize at 6 harmonic orders (28 directions), meaning that AP images can be reliably derived from HARDI sequences with more than or equal to 28 gradient directions.

>>>>

An alternative to FA is to use the generalized fractional anisotropy (GFA) image. Since FA shows low intensity in cross-fibre regions, it may adversely affect registration in those areas. GFA can properly characterize crossing-fibre intensities, and therefore may be a better candidate than FA. The GFA is computed from the SH coefficients of spherical harmonics based methods such as Q-ball, where the diffusion orientation distribution function (ODF) is denoted as  $\Psi(u)$ \cite{Tuch2004}, such that:
\begin{equation}
GFA = \sqrt{\frac{n \sum_{i=1}^{n}( \Psi(u_i) - \langle\Psi\rangle)^2} {(n-1) \sum_{i=1}^{n}\Psi(u_i)^2}}
\end{equation}

@@

Another promising candidate is the anisotropic power (AP) image \cite{DellAcqua2014}. AP have similar visual intensity to T$_1$, and can be computed from HARDI acquisitions without any acquisition modifications. Since AP are derived from the SH model, it is also capable of proper cross-fibre region characterization. AP images are derived from the even harmonic orders ($l$) of SH coefficients \cite{Descoteaux2006,Frank2002} and are defined as
\begin{equation}
AP = \sum_{l=2,4,...} \frac{1}{2l+1}\sum_{m=-l}^{l} \lvert a_{l,m}\lvert^2
\end{equation}
where l is the harmonic order, m is the l-th order coefficient index, a is the corresponding SH coefficient. The AP values are then normalized to non-negative scale by the function:
\begin{equation}
\ln( \frac{AP}{AP_{ref}})
\end{equation}
where $\text{AP}_{\text{ref}}$ is the normalization constant. The resulting image will closely match T$_1$ intensities, and can better capture contrasts in the gray matter structures. AP values stabilize at 6 harmonic orders (28 directions), meaning that AP images can be reliably derived from HARDI sequences with more than or equal to 28 gradient directions.

### 2018-04-26 contents/discussions.tex
Updates from ShareLaTeX

\item Imaging of the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography. Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibers could not be visualized as they enter into the pons. Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increasing the range of study of trigeminal-related pathology that can be studied with neuroimaging, such as MS-TN.

@@

\section{Tractography of the trigeminal nerve and brainstem fibers}

An important motivation for the use of tractography to study the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the fibers of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.

@@

Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter remains a limitation. The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure, therefore, poses a fundamental challenge for neuroimaging.
Techniques to improve visualization of white matter fibers consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map. This kind of image-based probabilistic tractography has proved useful for whole-brain white matter studies. However, it is constrained to the resolution of the diffusion acquisition, which can be insufficient for localized and small diameter pathways. Moreover, the algorithm depends on pre-generated tissue-priors, which excludes cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempt to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions where required) in the diffusion acquisition. The acquisition time of these sequences can be impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research.

The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), are a significant development. SH-based methods model the fiber orientation distribution as a decomposition of SH basis functions. This means that these methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with at least 50 gradient directions.

These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility. We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images. 2) it courses through simple (cistern) and complex (pons) fiber environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.

@@

\subsection{Delineating the brainstem fibers of the trigeminal nerve}

For successful tractography of the CN V, both DWI sequence and the algorithm choice are paramount. Given a DWI sequence, cisternal CN V could be readily reconstructed with Gaussian tensors. Unlike major white matter bundles, cranial nerves diameters are in the millimetre range, and therefore voxel sizes greater than 2 mm may result in partial volume averaging. Achieving 2 mm isovoxel on a 3T scanner requires a 32 channel head coil. Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable. For this reason, the acquisition in-plane resolution can be tuned to 1x1 mm, and slice thickness at 3mm. While the acquisition is not ideal, it sacrifices resolution only in the z-axis, however the tractography visual fidelity is improved. 
Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers. A number of methods have been devised, including probabilistic tractography (FSL)\cite{Behrens2007}, which generates image-based results that represent the propagation probability of a seed region. Neurosurgical visualization however requires three dimensional representation of the fibers, and therefore streamline-based tractography visualizations are of broader appeal. In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater than 50 gradient directions.

@@

\subsection{Tractography group registration}
Tractography is a powerful methodology that permits the visualization of white matter fibers in three dimensions. Its primary use, however, has been limited to case studies and small-scale anatomy comparative analysis. The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of larger groups.

@@

One limitation across the studies in this thesis was the choice of anisotropic voxels of the DWI sequences. This was limited by the clinical 3-channel GE 3T HDx MR scanner, which places constraints both the acquisition resolution and time. Commonly, DWI voxel resolution on a 32-channel 3T research scanner can obtain up to 2mm isovoxel resolution, on older 8-channel clinical scanners, the maximum isovoxel resolution is limited at about 2.6mm isovoxel. This is insufficient for the study of cranial nerves such as CN V. Our resolution is limited to 1x1x3 mm voxel dimensions, in order to retain in-plane resolution at the sacrifice of slice thickness. Despite this, certain regions such as the TPT could not be reliably reconstructed in all of the subjects. Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections which include the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution. Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher isovoxel resolution can be immediately processed as soon as they become available.

@@

\section{Future Directions}

@@

\section{Conclusions}

>>>>

\item Imaging of the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography. Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibres could not be visualized as they enter into the pons. Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increasing the range of study of trigeminal-related pathology that can be studied with neuroimaging, such as MS-TN.

@@

\section{Tractography of the trigeminal nerve and brainstem fibres}

An important motivation for the use of tractography to study the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the fibres of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.

@@

Diffusion tensor imaging revolutionized how we examine brain white matter fibres, yet resolving cross-fibres of the cerebral white matter remains a limitation. The brain white matter consists of complex crossing fibre crossings, and the inability for DTI to reliably image these structure, therefore, poses a fundamental challenge for neuroimaging.
Techniques to improve visualization of white matter fibres consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map. This kind of image-based probabilistic tractography has proved useful for whole-brain white matter studies. However, it is constrained to the resolution of the diffusion acquisition, which can be insufficient for localized and small diameter pathways. Moreover, the algorithm depends on pre-generated tissue-priors, which excludes cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempt to model the fibre orientation distribution by dramatically increasing the number of gradient directions (256--512 directions where required) in the diffusion acquisition. The acquisition time of these sequences can be impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research.

The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), are a significant development. SH-based methods model the fibre orientation distribution as a decomposition of SH basis functions. This means that these methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fibre crossings, also proved to provide good results with at least 50 gradient directions.

These algorithmic developments brought imaging of crossing-fibres using tractography to the realm of possibility. We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images. 2) it courses through simple (cistern) and complex (pons) fibre environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.

@@

\subsection{Delineating the brainstem fibres of the trigeminal nerve}

For successful tractography of the CN V, both DWI sequence and the algorithm choice are paramount. Given a DWI sequence, cisternal CN V could be readily reconstructed with Gaussian tensors. Unlike major white matter bundles, cranial nerves diameters are in the millimetre range, and therefore voxel sizes greater than 2 mm may result in partial volume averaging. Achieving 2 mm isovoxel on a 3T scanner requires a 32 channel head coil. Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable. For this reason, the acquisition in-plane resolution can be tuned to 1x1 mm, and slice thickness at 3mm. While the acquisition is not ideal, it sacrifices resolution only in the z-axis, however the tractography visual fidelity is improved. 
Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibres. A number of methods have been devised, including probabilistic tractography (FSL)\cite{Behrens2007}, which generates image-based results that represent the propagation probability of a seed region. Neurosurgical visualization however requires three dimensional representation of the fibres, and therefore streamline-based tractography visualizations are of broader appeal. In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater than 50 gradient directions.

@@

\subsection{Tractography group registration}
Tractography is a powerful methodology that permits the visualization of white matter fibres in three dimensions. Its primary use, however, has been limited to case studies and small-scale anatomy comparative analysis. The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of larger groups.

@@

One limitation across the studies in this thesis was the choice of anisotropic voxels of the DWI sequences. This was limited by the clinical 3-channel GE 3T HDx MR scanner, which places constraints both the acquisition resolution and time. Commonly, DWI voxel resolution on a 32-channel 3T research scanner can obtain up to 2mm isovoxel resolution, on older 8-channel clinical scanners, the maximum isovoxel resolution is limited at about 2.6mm isovoxel. This is insufficient for the study of cranial nerves such as CN V. Our resolution is limited to 1x1x3 mm voxel dimensions, in order to retain in-plane resolution at the sacrifice of slice thickness. Despite this, certain regions such as the TPT could not be reliably reconstructed in all of the subjects. Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections which include the claustrum, and medullary decussations of the spinal trigeminal fibres will remain out of reach of the current algorithms without substantial improvement in DWI resolution. Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher isovoxel resolution can be immediately processed as soon as they become available.

@@

\chapter{Conclusions}

\chapter{Future Directions}

@@

### 2018-04-26 contents/hypotheses.tex
Updates from ShareLaTeX

The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease. This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.

With an increasing number of subjects, single subject analysis becomes cumbersome. Critical methodological advances are necessary for the feasibility of the general aim. These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.

Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.  The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, in order to identify specific diffusivity pattern that differentiates between TN and non-TN. The roadmap and rationale for the specific studies are described below.

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

\subsection{Main Aim} 
Establish the feasibility of multi-tensor diffusion tractography to delineating the cisternal, root entry zone and brainstem segments of CN V and how diffusivity changes in these sub-regions correlate with pain in MS-TN.

@@

\subsection{Specific Hypothesis}
\begin{itemize}
    \item TN is known to relate to vascular compression of the cistern REZ, therefore its diffusivity signature will occur in the cisternal region.
    \item MS-TN is related to brainstem MS lesions. Therefore CN V segments close to the lesions will show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not. 
    \item As a demyelinating disease, the pattern of change for MS-TN diffusivity signatures will reflects demyelination, including increase in RD, and decrease in FA. 
\end{itemize}

@@

\subsection{Main Aim}
The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population. The validations are performed  at a scale that would be prohibitive with manual methods. The comparisons determine the best tractography methods to use for different types of neuroanatomy.

\subsection{Specific Aim}
\begin{itemize}
    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD. 
    \item Use the above techniques to delineate a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
    \item Validate normalized overlapping scores (NOS) by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item DTI is known to not able to delineate cross-fibers regions. If the group tractography works consistently, then its output will reflect such known algorithmic limitations, and DTI group tracts will not be able to perform well on more complex anatomies. The grouped tracts will however show consistent regional anatomical overlap across individuals.
    \item Specific tractography algorithms will be preferentially suited for the visualization of specific areas in the brain. For examples, deterministic tractography will preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
    \item NOS score will correlate with expert ratings, and show improved rating consistency over human rater variability.
\end{itemize}

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The developed methodology is applied to the study of the trigeminal sensory pathway. The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study. This study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. Along-the-tract analysis are carried out by dividing each segment into 30 subdivisions. Following this, we use machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.

>>>>

This thesis consists of two components: 1) to develop and validate methodology to study specific white matter in individuals and in groups of individuals, and 2) to apply the developed methods to gain insight into the changes in the trigeminal system in TN. The methodology developed for group analysis lays the foundation for future clinical applications.

The general goal was to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease. This necessitated detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.

With an increasing number of subjects, single subject analysis can become cumbersome. Critical methodological advances were necessary for the feasibility of the general aim. These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.

Lastly, the methodological advances of this thesis were applied to the large-scale study of TN versus healthy control.  The considerable amount of data output from the methodology  enabled the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, and in order to identify specific diffusivity pattern differentiating TN from healthy controls. The roadmap and rationale for the specific studies are described below.

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibres in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

\subsection{Main Aim} 
Establish the feasibility of multi-tensor diffusion tractography to delineate the cisternal, root entry zone and brainstem segments of CN V and how diffusivity alterations in these sub-regions correlate with pain in MS-TN.

@@

\subsection{Specific Hypothesis}
\begin{itemize}
    \item MS-TN is pathophysiologcially related to brainstem MS lesions. Therefore CN V segments close to these lesions will demonstrate diffusivity alterations that correlate with pain laterality.
    \item As a demyelinating disease, the pattern of change for MS-TN diffusivity signatures will reflects demyelination, including increase in RD, and decrease in FA. 
\end{itemize}

@@

\subsection{Main Objective}
The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population. The validations are performed  at a scale that would be prohibitive with manual methods. The comparisons determine the best tractography methods to use for different types of neuroanatomy.

\subsection{Specific Objectives}
\begin{itemize}
    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD.

    \item Use the above techniques to delineate a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.

    \item Determine if specific tractography algorithms are preferentially suited to delineate specific pathways in the brain. For example, deterministic tractography may show more specificity in delineating cranial nerves, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.

    \item Validate normalized overlapping scores (NOS) by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts. Demonstrate that NOS score will correlate with expert ratings, and show improved rating consistency over human rater variability.

\end{itemize}

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The methodology developed in the previous study was applied to the study of the trigeminal sensory pathway. The aim of this study was to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway was divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. Along-the-tract analysis were carried out by dividing each segment into 30 subdivisions. Following this, we used machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.

### 2018-04-26 contents/paper1.tex
Updates from ShareLaTeX

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure1.png}
\caption[Figure shows the MR images of a typical MS-TN patient at the level of pons.]{Figure shows the MR images of a typical MS-TN patient at the level of pons. The presence of MS plaque in the CNS at the level of trigeminal nerve is characteristic of MS-TN (panel A). Panel B shows the expected course of the trigeminal nerve (CN V) as it synapses onto the principal trigeminal and motor trigeminal nuclei. Note that the exact anatomical relationship between the MS plaque and CN V in this instance cannot be clearly discerned in T2 due to the lack of contrast between the nerve and the surrounding tissue (panels A,B,C). Panels C and D show the placements of the ROIs along the CN V nerve bilaterally in order to measure diffusivity statistics, overlaid with the axial anatomical/ DTI image. The colors of the DTI image are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. Diffusion statistics were measured from four groups of regions: cisternal segment, root entry zone (REZ), pontine, and peri-lesional, as illustrated. Analogous regions in TN patients and controls were similarly placed. Care was taken to differentiate pontine and peri- lesional ROIs in order to measure lesioned and unlesioned regions.}
\centering
\label{fig:MSfigure1}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure2.png}
\caption[Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views.]{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views. The reconstructed tracts are overlaid onto the axial DTI/anatomical images. The colors of the underlying scan are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. The colors of the tractography fibers represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner. Panel A shows that SDT tractography delineated the cerebellar peduncle fibers, but cannot distinguish the brainstem trigeminal fibers (yellow arrows). Red arrows denote the starting point of tract generation for both SDT and MTT methods. Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei.}
\centering
\label{fig:MSfigure2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure3.png}
\caption[Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method.]{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method. Images are shown in superior axial view. Panel A shows the MS-TN tractography delineations. There is focal decrease in CN V FA in the region of the MS lesion (yellow arrow). Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment. Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of color scales and figure annotations.}
\centering
\label{fig:MSfigure3}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure4.png}
\caption[Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.]{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls. Colored squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate. The (+/-) sign indicate symptomatic (+) and asymptomatic (-) sides of the subject group. The matrix should be read as ‘Y axis in relation to X axis’—for example, A red square indicates that the vertical axis measure (e.g. TN+, the affected side in TN patients) is significantly higher than that of the horizontal axis measure (e.g. MS-TN+, the affected side in MS-TN patients); blue indicates that the measure is lower. The number within the square indicates relative estimated differences. Bottom row: Box-plots of the diffusivity distributions. Note that the control group does not have pain lateralities, therefore ipsilateral indicate the right side in the control group in this figure. The mean of the left-right diffusivities in controls were used for comparisons against other groups. The analysis highlights TN+ specific diffusivity differences in the Cisternal segment of the trigeminal nerve. Specifically TN+ has significantly higher FA than all the other groups, and that there are significant differences between TN+ and TN-. The points above and below the box plots are outliers. Significances are not indicated in the plot, for pair-wise significance, please refer to the estimate matrices above row.}
\centering
\label{fig:MSfigure4}
\end{figure}

@@

Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus. While AD has been associated with neurological disruptions in MS \cite{Budde2009,Kim2006}, the present study showed a lack of change in AD. This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain. Disease severity was in fact variable in this group. Future studies will be able to address this point in more detail, by taking into account other MS manifestations in addition to pain, and correlation with diffusion metrics. Lower REZ FA in MS-TN compared to controls suggests that global neural tissue diffusivity in MS-TN is significantly different from that of healthy controls. Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups. However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibers in the pontine region may introduce confounds into these correlations. Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.

>>>>

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure1.png}
\caption[Figure shows the MR images of a typical MS-TN patient at the level of pons.]{Figure shows the MR images of a typical MS-TN patient at the level of pons. The presence of MS plaque in the CNS at the level of trigeminal nerve is characteristic of MS-TN (panel A). Panel B shows the expected course of the trigeminal nerve (CN V) as it synapses onto the principal trigeminal and motor trigeminal nuclei. Note that the exact anatomical relationship between the MS plaque and CN V in this instance cannot be clearly discerned in T2 due to the lack of contrast between the nerve and the surrounding tissue (panels A,B,C). Panels C and D show the placements of the ROIs along the CN V nerve bilaterally in order to measure diffusivity statistics, overlaid with the axial anatomical/ DTI image. The colours of the DTI image are rendered as colour-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. Diffusion statistics were measured from four groups of regions: cisternal segment, root entry zone (REZ), pontine, and peri-lesional, as illustrated. Analogous regions in TN patients and controls were similarly placed. Care was taken to differentiate pontine and peri- lesional ROIs in order to measure lesioned and unlesioned regions.}
\centering
\label{fig:MSfigure1}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure2.png}
\caption[Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views.]{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views. The reconstructed tracts are overlaid onto the axial DTI/anatomical images. The colours of the underlying scan are rendered as colour-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. The colours of the tractography fibres represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner. Panel A shows that SDT tractography delineated the cerebellar peduncle fibres, but cannot distinguish the brainstem trigeminal fibres (yellow arrows). Red arrows denote the starting point of tract generation for both SDT and MTT methods. Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei.}
\centering
\label{fig:MSfigure2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure3.png}
\caption[Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method.]{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method. Images are shown in superior axial view. Panel A shows the MS-TN tractography delineations. There is focal decrease in CN V FA in the region of the MS lesion (yellow arrow). Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment. Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of colour scales and figure annotations.}
\centering
\label{fig:MSfigure3}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure4.png}
\caption[Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.]{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls. coloured squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate. The (+/-) sign indicate symptomatic (+) and asymptomatic (-) sides of the subject group. The matrix should be read as ‘Y axis in relation to X axis’—for example, A red square indicates that the vertical axis measure (e.g. TN+, the affected side in TN patients) is significantly higher than that of the horizontal axis measure (e.g. MS-TN+, the affected side in MS-TN patients); blue indicates that the measure is lower. The number within the square indicates relative estimated differences. Bottom row: Box-plots of the diffusivity distributions. Note that the control group does not have pain lateralities, therefore ipsilateral indicate the right side in the control group in this figure. The mean of the left-right diffusivities in controls were used for comparisons against other groups. The analysis highlights TN+ specific diffusivity differences in the Cisternal segment of the trigeminal nerve. Specifically TN+ has significantly higher FA than all the other groups, and that there are significant differences between TN+ and TN-. The points above and below the box plots are outliers. Significances are not indicated in the plot, for pair-wise significance, please refer to the estimate matrices above row.}
\centering
\label{fig:MSfigure4}
\end{figure}

@@

Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus. While AD has been associated with neurological disruptions in MS \cite{Budde2009,Kim2006}, the present study showed a lack of change in AD. This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain. Disease severity was in fact variable in this group. Future studies will be able to address this point in more detail, by taking into account other MS manifestations in addition to pain, and correlation with diffusion metrics. Lower REZ FA in MS-TN compared to controls suggests that global neural tissue diffusivity in MS-TN is significantly different from that of healthy controls. Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups. However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibres in the pontine region may introduce confounds into these correlations. Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.
